---
title: "Permutation"
author: Astrid Deschenes, Pascal Belleau and Arnaud Droit 
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{Permutation Positioning}
  %\VignettePackage{methylInheritance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
library(knitr)
```

<br />
**Package**: `r Rpackage("methylInheritance")`<br />
**Authors**: `r packageDescription("methylInheritance")[["Author"]]`<br />
**Version**: `r packageDescription("methylInheritance")$Version`<br />
**Compiled date**: `r Sys.Date()`<br />
**License**: `r packageDescription("methylInheritance")[["License"]]`<br />

# Licensing 

The **methylInheritance** package and the underlying **methylInheritance** code 
are distributed under the Artistic license 2.0. You are free to use and 
redistribute this software. 


# Loading methylInheritance package

As with any R package, the `r Rpackage("methylInheritance")` package should 
first be loaded with the following command:

```{r loadingPackage, warning=FALSE, message=FALSE}
library(methylInheritance)
```


```{r loadingFiles, warning=FALSE, message=FALSE}
F1_control_01 <- system.file("extdata",
    "F1_control_01_Cpg.txt", package = "methylInheritance")
F1_control_02 <- system.file("extdata",
    "F1_control_02_Cpg.txt", package = "methylInheritance")
F1_control_03 <- system.file("extdata",
    "F1_control_03_Cpg.txt", package = "methylInheritance")
F1_case_01 <- system.file("extdata",
    "F1_case_01_Cpg.txt", package = "methylInheritance")
F1_case_02 <- system.file("extdata",
    "F1_case_02_Cpg.txt", package = "methylInheritance")
F1_case_03 <- system.file("extdata",
    "F1_case_03_Cpg.txt", package = "methylInheritance")
F2_control_01 <- system.file("extdata",
    "F2_control_01_Cpg.txt", package = "methylInheritance")
F2_control_02 <- system.file("extdata",
    "F2_control_02_Cpg.txt", package = "methylInheritance")
F2_control_03 <- system.file("extdata",
    "F2_control_03_Cpg.txt", package = "methylInheritance")
F2_case_01 <- system.file("extdata",
    "F2_case_01_Cpg.txt", package = "methylInheritance")
F2_case_02 <- system.file("extdata",
    "F2_case_02_Cpg.txt", package = "methylInheritance")
F2_case_03 <- system.file("extdata",
    "F2_case_03_Cpg.txt", package = "methylInheritance")
F3_control_01 <- system.file("extdata",
    "F3_control_01_Cpg.txt", package = "methylInheritance")
F3_control_02 <- system.file("extdata",
    "F3_control_02_Cpg.txt", package = "methylInheritance")
F3_control_03 <- system.file("extdata",
    "F3_control_03_Cpg.txt", package = "methylInheritance")
F3_case_01 <- system.file("extdata",
    "F3_case_01_Cpg.txt", package = "methylInheritance")
F3_case_02 <- system.file("extdata",
    "F3_case_02_Cpg.txt", package = "methylInheritance")
F3_case_03 <- system.file("extdata",
    "F3_case_03_Cpg.txt", package = "methylInheritance")

F3_case_02
```

```{r}

conditions <- c(0,0,0,1,1,1)

dataNames <- c("F1A","F1B","F1C","F1D","F1E","F1F",
               "F2A","F2B","F2C","F2D","F2E","F2F",
               "F3A","F3B","F3C","F3D","F3E","F3F")

allFilesByGeneration <- list(F1=list(F1_control_01, F1_control_02, F1_control_03, F1_case_01, F1_case_02, F1_case_03), 
                             F2=list(F2_control_01, F2_control_02, F2_control_03, F2_case_01, F2_case_02, F2_case_03), 
                             F3=list(F3_control_01, F3_control_02, F3_control_03, F3_case_01, F3_case_02, F3_case_03))
conditionsByGeneration <- list(F1=conditions, F2=conditions, F3=conditions)
output_dir <- "test"
genomeVersion <- "rnor5"
nbrPermutations <- 2
nbrCores <- 1
nbrCoresDiffMeth <- 1
doingSites <- TRUE
doingTiles <- FALSE

runPermutation(allFilesByGeneration, conditionsByGeneration,
                                output_dir = output_dir,
                                genomeVersion = genomeVersion,
                                nbrPermutations = nbrPermutations,
                                nbrCores = nbrCores,
                                nbrCoresDiffMeth = nbrCoresDiffMeth,
                                doingSites = doingSites,
                                doingTiles = doingTiles,
                                minReads = 10,
                                minMethDiff = 20,
                                qvalue = 0.01,
                                maxPercReads = 99.9,
                                destrand = FALSE,
                                tileSize = 1000,
                                stepSize = 1000,
                                vSeed = -1)

```

