---
title: "Conserved Differentially Methylated Elements from One Generation to the Next: Inheritance versus Randomness"
author: Astrid Deschenes, Pascal Belleau and Arnaud Droit 
output:
  BiocStyle::html_document:
    toc: true
bibliography: biblio.bibtex
vignette: >
  %\VignetteIndexEntry{Permutation Positioning}
  %\VignettePackage{methylInheritance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
library(knitr)
```

<br />
**Package**: `r Rpackage("methylInheritance")`<br />
**Authors**: `r packageDescription("methylInheritance")[["Author"]]`<br />
**Version**: `r packageDescription("methylInheritance")$Version`<br />
**Compiled date**: `r Sys.Date()`<br />
**License**: `r packageDescription("methylInheritance")[["License"]]`<br />


# Licensing 

The **methylInheritance** package and the underlying **methylInheritance** code 
are distributed under the Artistic license 2.0. You are free to use and 
redistribute this software. 


# Introduction

DNA methylation plays an important role in the biology of tissue development 
and diseases. High-throughput sequencing techniques enable genome-wide 
detection of differentially methylated elements (DME), commonly sites (DMS) or 
regions (DMR). The analysis of treatment effects on DNA methylation, from 
one generation to the next (inter-generational) and across generations that 
were not exposed to the initial environment (trans-generational) represent 
complex designs. Due to software design, the detection of DME is usually 
made on each generation separately. However, the common DME between 
generations due to randomness is not negligible when the number of DME 
detected in each generation is high. To judge the effect on DME that is 
inherited from a treatment in previous generation, the observed number of 
conserved DME must be compared to the randomly expected number. 

We present a permutation analysis, based on Monte Carlo sampling, aim to infer 
a relation between the number of conserved DME from one generation to the next 
to the inheritance effect of treatment and to dismiss stochastic effect.  It 
is used as a robust alternative to inference based on parametric assumptions.

The `r Rpackage("methylInheritance")` package can perform a permutation 
analysis on both differentially methylated sites (DMS) and differentially 
methylated tiles (DMT) using `r Biocpkg("methylKit")` package.


# Loading methylInheritance package

As with any R package, the `r Rpackage("methylInheritance")` package should 
first be loaded with the following command:

```{r loadingPackage, warning=FALSE, message=FALSE}
library(methylInheritance)
```


# Description of the permutation analysis

The permutation analysis is a statistical significance test in which 
the distribution of the test statistic under the null hypothesis is obtained 
by calculating the values of the test statistic under rearrangements of 
the labels on the observed data points. The rearrangements of the labels is 
done through repeated random sampling [@Legendre1998, pp. 142-157]. 

**Null Hypothesis**: The number of conserved DME correspond to a number that 
can be obtain through a randomness analysis. 

**Alternative Hypothesis**: The number of conserved DME do not correspond to a 
number that can be obtain through a randomness analysis.

A typical **methylInheritance** analysis consists of the following steps:

1. Process to a differentially methylation analysis on each generation 
separatly using real dataset with `r Rpackage("methylKit")`.
2. Calculate the number of conserved differentially methylated elements 
between two consecutive generations (F1 and F2, F2 and F3, etc..).  The number
of conserved differentially methylated elements is also calculated for three
or more consercutive generations, always starting with the first generation 
(F1 and F2 and F3, F1 and F2 and F3 and F4, etc..).
Those results are considered the reference values.
3. Fix a threshold (conventionally 0.05) that is used as a cutoff between the 
null and alternative hypothesis. 
4. Process to a differentially methylation analysis on each permutated dataset.
Each generation is analysed separatly using `r Rpackage("methylKit")` package.
5. Calculate the significant level for each consecutive subset of generations. 
The significant level is defined as the percentage of results equal or higher 
than the reference values. The reference vaules are added to the analysis so 
that it becomes impossible for the test to conclude that no value are 
as extreme as, or more extreme than the reference values.

All those steps have been encoded in the 
**methylInheritance** package. 

# Case study 

## The dataset

A dataset containing methylation data (6 cases and 6 controls) over three 
generations has been generated using the 
`r Rpackage("methylInheritanceSim")` package.

```{r caseStudy01, warning=FALSE, message=FALSE, collapse=TRUE}
## Load dataset containing information over three generations
data(demoForTransgenerationalAnalysis)

## The length of the dataset corresponds to the number of generation
## The generations are stored in order (first entry = first generation, 
## second entry = second generation, etc..)
length(demoForTransgenerationalAnalysis)


## All samples related to one generations are contained in a S4 methylRawList object.
## The methylRawList object contains two Slots:
## 1- treatment: a numeric vector denoting controls and cases.
## 2- .Data: a list of methylRaw objects. Each object stores the raw mehylation 
## data of one sample.


## A section of the methylRaw object containing the information of the first 
## sample from the second generation 
head(demoForTransgenerationalAnalysis[[2]][[1]]) 

## The same information could also be obtained using the .Data slot
## head(demoForTransgenerationalAnalysis[[2]]@.Data[[1]]) 

## The treatment vector for each generation
## The number of treatments and controls is the same in each generation
## However, it could also be different.
slot(demoForTransgenerationalAnalysis[[1]], "treatment")
slot(demoForTransgenerationalAnalysis[[2]], "treatment")
slot(demoForTransgenerationalAnalysis[[3]], "treatment")
```


## Observation analysis 

The observation analysis can be run on all generations using the
*runObservationUsingMehtylKitInfo()* function.

The observation results can optionally be stored in a RDS file. To store the
results into a RDS file, the *outputDir* parameter must be given a directory
path.

```{r caseStudy02, warning=FALSE, message=FALSE, collapse=TRUE, cache=TRUE}
## The observation analysis is only done on differentially methylated sites
observationResult <- runObservationUsingMethylKitInfo(methylKitInfo = 
                                        demoForTransgenerationalAnalysis, 
                                type = "sites",     # Only sites
                                outputDir = NULL,   # RDS files only created when a path is given
                                nbrCores = 1,
                                minReads = 10,
                                minMethDiff = 10,
                                qvalue = 0.01,
                                vSeed = 2101)

observationResult
```


## Permutation analysis 

The permutation analysis can be run on all generations using the
*runPermutationUsingMehtylKitInfo()* function.

The observation and the permutation analysis can be run together by 
setting the *runObservationAnalysis = TRUE* in the
*runPermutationUsingMehtylKitInfo()* function.

All permutations can also be, optionally, saved in RDS files. To store the
results into RDS files, the *outputDir* parameter must be given a directory
path.

At last, the *runPermutationUsingRDSFile()* function also executes a
permutation analysis but insteed of passing a methylKit object to it, the
path to a RDS file that contains the methylKit object must be used.

```{r caseStudy03, warning=FALSE, message=FALSE, collapse=TRUE, cache=TRUE}

## The permutation analysis is only done on differentially methylated sites
permutationResult <- runPermutationUsingMethylKitInfo(methylKitInfo = 
                                        demoForTransgenerationalAnalysis, 
                                type = "sites",     # Only sites
                                outputDir = NULL,   # RDS files created
                                runObservationAnalysis = FALSE,
                                nbrCores = 1,
                                nbrPermutations = 3,
                                minReads = 10,
                                minMethDiff = 10,
                                qvalue = 0.01,
                                vSeed = 2101)

permutationResult
```


## Merging observation and permutation analysis

The observation and permutation results can be merged using the
*mergePermutationAndObservation()* function.

```{r caseStudy04, warning=FALSE, message=FALSE, collapse=TRUE, cache=TRUE}
## Merge observation and permutation results
allResults <- mergePermutationAndObservation(permutationResults = permutationResult,
                                    observationResults = observationResult)
```

When observation and permutation analysis have been run together using the
*runPermutationUsingMehtylKitInfo()* function, this step can be skipped.

## Extract a specific analysis

The *runPermutationUsingMehtylKitInfo()* and *runObservationUsingMehtylKitInfo()* functions 
calculate the number of conserved differentially methylated elements 
between two consecutive generations (F1 and F2, F2 and F3, etc..).  They also calculate 
the number of conserved differentially methylated elements is also calculated for three
or more consercutive generations, always starting with the first generation 
(F1 and F2 and F3, F1 and F2 and F3 and F4, etc..).

A specific analysis can be extracted from the results using *extractInfo()* function.

The *type* parameter can be set to extract one of those elements:

* *sites*: differentially methylated sites
* *tiles*: differentially methylated tiles

The *inter* parameter can be set to extract one of those analysis type:

* *i2*: the analysis between two consecutive generations (F1 and F2, F2 and F3, etc..)
* *iAll*: the analysis between three or more generations (F1 and F2 and F3, F1 and F2 and F3 and F4, etc..)

```{r caseStudy05, warning=FALSE, message=FALSE, collapse=TRUE}
## Differentially conserved sites between F1 and F2 generations
F1_and_F2_results <- extractInfo(allResults = allResults, type = "sites", inter = "i2", position = 1)

head(F1_and_F2_results)
```

## Significant level and visual representation

The permutation analysis has been run on the *demoForTransgenerationalAnalysis* with 
1000 permutations (*nbrPermutation = 1000*). The results of those permutations will be used
to generate the singificant levels and the visual representations.

```{r caseStudyLoad, warning=FALSE, message=FALSE, cache=TRUE,  echo = FALSE, cache=TRUE}
demoResults <- readRDS(dir(system.file("extdata", package = "methylInheritance"),
                       pattern = "resultsForTransgenerationalAnalysis.RDS", full.names = TRUE))
```

```{r caseStudy06, warning=FALSE, message=FALSE, collapse=TRUE, cache=TRUE}
## Differentially conserved sites between F1 and F2 generations
F1_and_F2 <- extractInfo(allResults = demoResults, type = "sites", inter = "i2", position = 1)
## Differentially conserved sites between F2 and F3 generations
F2_and_F3 <- extractInfo(allResults = demoResults, type = "sites", inter = "i2", position = 2)
## Differentially conserved sites between F1 and F2 and F3 generations
F2_and_F3 <- extractInfo(allResults = demoResults, type = "sites", inter = "iAll", position = 1)
```

```{r caseStudy07, warning=FALSE, message=FALSE, collapse=TRUE}
## Show graph and significant level for differentially conserved sites between F1 and F2 
output <- plotGraph(F1_and_F2)
```

# References


